---
title: "Protein Overlap"
output:
  html_document:
    echo: no
  pdf_document: default
---

```{r set_options, echo=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(echo = FALSE)

```


Protein overlap among results of 6600_64w of:

1. Spectronaut  
1. PeakView  
1. Skyline  
1. OpenSWATH  
1. DIAumpire  

Peptide identifications of other software tools are almost completely covered by Spectronaut. 

```{r loadlibs}

loadLibrary <- function(x) 
{
  if(!require(x, character.only=T, quietly = T, warn.conflicts=F)) 
  { 
    install.packages(x); library(x, character.only=T, warn.conflicts=F) 
  } 
}

loadLibrary("VennDiagram")
loadLibrary("tidyr")
loadLibrary("dplyr")
loadLibrary("ggplot2")

```


```{r loadFiles}

diaump.file <- "/Users/napedro/github_prj/swath-evaluation-manuscript/benchmarks/draft.v2/dataanalysis_TOF6600_64w/input/DIAumpire_r1.tsv"
spectronaut.file <- "/Users/napedro/github_prj/swath-evaluation-manuscript/benchmarks/draft.v2/dataanalysis_TOF6600_64w/input/Spectronaut_r2.tsv"
skyline.file <- "/Users/napedro/github_prj/swath-evaluation-manuscript/benchmarks/draft.v2/dataanalysis_TOF6600_64w/input/Skyline_r2.tsv"
peakview.file <- "/Users/napedro/github_prj/swath-evaluation-manuscript/benchmarks/draft.v2/dataanalysis_TOF6600_64w/input/PeakView_r2.tsv"
openswath.file <- "/Users/napedro/github_prj/swath-evaluation-manuscript/benchmarks/draft.v2/dataanalysis_TOF6600_64w/input/OpenSWATH_r2.tsv"

spectronaut.r1.file <- "/Users/napedro/github_prj/swath-evaluation-manuscript/benchmarks/draft.v2/dataanalysis_TOF6600_64w/input/Spectronaut_r1.tsv"
skyline.r1.file <- "/Users/napedro/github_prj/swath-evaluation-manuscript/benchmarks/draft.v2/dataanalysis_TOF6600_64w/input/Skyline_r1.tsv"
peakview.r1.file <- "/Users/napedro/github_prj/swath-evaluation-manuscript/benchmarks/draft.v2/dataanalysis_TOF6600_64w/input/PeakView_r1.tsv"
openswath.r1.file <- "/Users/napedro/github_prj/swath-evaluation-manuscript/benchmarks/draft.v2/dataanalysis_TOF6600_64w/input/OpenSWATH_r1.tsv"


diaump <- read.table(diaump.file, header=T, sep="\t", stringsAsFactors=F)
spectronaut <- read.table(spectronaut.file, header=T, sep="\t", stringsAsFactors=F)
skyline <- read.table(skyline.file, header=T, sep="\t", stringsAsFactors=F)
peakview <- read.table(peakview.file, header=T, sep="\t", stringsAsFactors=F)
openswath <- read.table(openswath.file, header=T, sep="\t", stringsAsFactors=F)

spectronaut.r1 <- read.table(spectronaut.r1.file, header=T, sep="\t", stringsAsFactors=F)
skyline.r1 <- read.table(skyline.r1.file, header=T, sep="\t", stringsAsFactors=F)
peakview.r1 <- read.table(peakview.r1.file, header=T, sep="\t", stringsAsFactors=F)
openswath.r1 <- read.table(openswath.r1.file, header=T, sep="\t", stringsAsFactors=F)

```

Parse protein IDs from each software and ungroup them. There is not an easy way to deal with protein groups: they are not equivalent from one software to the other. As a rough approach, we can consider _all_ proteins of a group as identified. This is going to be a pesimistic approach for the protein overlap among software tools.

```{r parseIDs}

spectronaut <- spectronaut %>% 
    mutate( proteinID = substring(proteinID, first=4, last=nchar(proteinID))) %>%
    mutate( proteins = strsplit(proteinID, split="/")) %>%
    unnest(proteins) %>%
    mutate( protein = gsub("\\|", "", regmatches(proteins, gregexpr("\\|.*?\\|", proteins))) ) %>%
    select(-proteins) %>% 
    distinct(protein) %>% 
    group_by(protein) %>%
    mutate( meanIntensity = mean(c(A1, A2, A3, B1, B2, B3), na.rm=T))

spectronaut.id <- spectronaut %>%
    select(protein) %>%
    mutate(Spectronaut = TRUE)

diaump <- diaump %>% 
    mutate( proteins = strsplit(proteinID, split=";")) %>%
    unnest(proteins) %>%
    mutate( protein = gsub("\\|", "", regmatches(proteins, gregexpr("\\|.*?\\|", proteins))) ) %>%
    select(-proteins) %>% 
    distinct(protein) %>% 
    group_by(protein) %>%
    mutate( meanIntensity = mean(c(A1, A2, A3, B1, B2, B3), na.rm=T))

diaump.id <- diaump %>%
    select(protein) %>%
    mutate(DIAumpire = TRUE)

skyline <- skyline %>% 
    mutate( proteins = strsplit(proteinID, split="/")) %>%
    unnest(proteins) %>%
    filter(nchar(proteins) > 2) %>%
    mutate( protein = gsub("\\|", "", regmatches(proteins, gregexpr("\\|.*?\\|", proteins))) ) %>%
    select(-proteins) %>% 
    distinct(protein) %>% 
    group_by(protein) %>%
    mutate( meanIntensity = mean(c(A1, A2, A3, B1, B2, B3), na.rm=T))

skyline.id <- skyline %>%
    select(protein) %>%
    mutate(Skyline = TRUE)

peakview <- peakview %>% 
    mutate( proteins = strsplit(proteinID, split="/")) %>%
    unnest(proteins) %>%
    filter(nchar(proteins) > 2) %>%
    mutate( protein = gsub("\\|", "", regmatches(proteins, gregexpr("\\|.*?\\|", proteins))) ) %>%
    select(-proteins) %>% 
    distinct(protein) %>% 
    group_by(protein) %>%
    mutate( meanIntensity = mean(c(A1, A2, A3, B1, B2, B3), na.rm=T))

peakview.id <- peakview %>%
    select(protein) %>%
    mutate(PeakView = TRUE)


openswath <- openswath %>% 
    mutate( proteins = strsplit(proteinID, split="/")) %>%
    unnest(proteins) %>%
    filter(nchar(proteins) > 2) %>%
    mutate( protein = gsub("\\|", "", regmatches(proteins, gregexpr("\\|.*?\\|", proteins))) ) %>%
    select(-proteins) %>% 
    distinct(protein) %>% 
    group_by(protein) %>%
    mutate( meanIntensity = mean(c(A1, A2, A3, B1, B2, B3), na.rm=T))

openswath.id <- openswath %>%
    select(protein) %>%
    mutate(OpenSWATH = TRUE)


proteinoverlap <- merge(spectronaut.id, diaump.id, by="protein", all=T)
proteinoverlap <- merge(proteinoverlap, peakview.id, by="protein", all=T)
proteinoverlap <- merge(proteinoverlap, skyline.id, by="protein", all=T)
proteinoverlap <- merge(proteinoverlap, openswath.id, by="protein", all=T)

proteinoverlap[is.na(proteinoverlap)] <- FALSE


# for the iteration 1

spectronaut.r1 <- spectronaut.r1 %>% 
    mutate( proteinID = substring(proteinID, first=4, last=nchar(proteinID))) %>%
    mutate( proteins = strsplit(proteinID, split="/")) %>%
    unnest(proteins) %>%
    mutate( protein = gsub("\\|", "", regmatches(proteins, gregexpr("\\|.*?\\|", proteins))) ) %>%
    select(-proteins) %>% 
    distinct(protein) %>% 
    group_by(protein) %>%
    mutate( meanIntensity = mean(c(A1, A2, A3, B1, B2, B3), na.rm=T))

spectronaut.r1.id <- spectronaut.r1 %>%
    select(protein) %>%
    mutate(Spectronaut = TRUE)

skyline.r1 <- skyline.r1 %>% 
    mutate( proteins = strsplit(proteinID, split="/")) %>%
    unnest(proteins) %>%
    filter(nchar(proteins) > 2) %>%
    mutate( protein = gsub("\\|", "", regmatches(proteins, gregexpr("\\|.*?\\|", proteins))) ) %>%
    select(-proteins) %>% 
    distinct(protein) %>% 
    group_by(protein) %>%
    mutate( meanIntensity = mean(c(A1, A2, A3, B1, B2, B3), na.rm=T))

skyline.r1.id <- skyline.r1 %>%
    select(protein) %>%
    mutate(Skyline = TRUE)

peakview.r1 <- peakview.r1 %>% 
    mutate( proteins = strsplit(proteinID, split="/")) %>%
    unnest(proteins) %>%
    filter(nchar(proteins) > 2) %>%
    mutate( protein = gsub("\\|", "", regmatches(proteins, gregexpr("\\|.*?\\|", proteins))) ) %>%
    select(-proteins) %>% 
    distinct(protein) %>% 
    group_by(protein) %>%
    mutate( meanIntensity = mean(c(A1, A2, A3, B1, B2, B3), na.rm=T))

peakview.r1.id <- peakview.r1 %>%
    select(protein) %>%
    mutate(PeakView = TRUE)


openswath.r1 <- openswath.r1 %>% 
    mutate( proteins = strsplit(proteinID, split="/")) %>%
    unnest(proteins) %>%
    filter(nchar(proteins) > 2) %>%
    mutate( protein = gsub("\\|", "", regmatches(proteins, gregexpr("\\|.*?\\|", proteins))) ) %>%
    select(-proteins) %>% 
    distinct(protein) %>% 
    group_by(protein) %>%
    mutate( meanIntensity = mean(c(A1, A2, A3, B1, B2, B3), na.rm=T))

openswath.r1.id <- openswath.r1 %>%
    select(protein) %>%
    mutate(OpenSWATH = TRUE)


proteinoverlap.r1 <- merge(spectronaut.r1.id, diaump.id,       by="protein", all=T)
proteinoverlap.r1 <- merge(proteinoverlap   , peakview.r1.id,  by="protein", all=T)
proteinoverlap.r1 <- merge(proteinoverlap   , skyline.r1.id,   by="protein", all=T)
proteinoverlap.r1 <- merge(proteinoverlap   , openswath.r1.id, by="protein", all=T)

proteinoverlap.r1[is.na(proteinoverlap.r1)] <- FALSE



#List of proteins present in all tools
commonproteins <- proteinoverlap %>% (filter(Spectronaut == T & 
                                            PeakView    == T & 
                                            OpenSWATH   == T & 
                                            Skyline     == T & 
                                            DIAumpire   == T))

commonproteins.list <- commonproteins$protein
write.table(commonproteins.list, file="list_common_proteins.txt", sep="\t", row.names = F, col.names = F, quote = F)

commonproteins.r1 <- proteinoverlap.r1 %>% (filter(Spectronaut == T & 
                                            PeakView    == T & 
                                            OpenSWATH   == T & 
                                            Skyline     == T & 
                                            DIAumpire   == T))

commonproteins.r1.list <- commonproteins.r1$protein
write.table(commonproteins.r1.list, file="list_common_proteins.txt", sep="\t", row.names = F, col.names = F, quote = F)


```


We are interested on how many proteins of all library-based tools are not covered by Spectronaut:

```{r}
l <- venn.diagram(
                x = list(
                    Spectronaut = unlist(proteinoverlap %>% filter(Spectronaut == T) %>% select(protein)),
                    PeakView = unlist(proteinoverlap %>% filter(PeakView == T) %>% select(protein)),
                    OpenSWATH = unlist(proteinoverlap %>% filter(OpenSWATH == T) %>% select(protein)),
                    Skyline = unlist(proteinoverlap %>% filter(Skyline == T) %>% select(protein))
                    ),
                filename = "Figure5C.tiff",
                fontfamily="Arial",
                fill=c("yellow", "red", "green", "cyan"),
                cex=1.5, cat.cex=1.1, cat.dist=0.07
                #,category.names = rep("", 4)
                )

l <- venn.diagram(
                x = list(
                    Spectronaut = unlist(proteinoverlap.r1 %>% filter(Spectronaut == T) %>% select(protein)),
                    PeakView = unlist(proteinoverlap.r1 %>% filter(PeakView == T) %>% select(protein)),
                    OpenSWATH = unlist(proteinoverlap.r1 %>% filter(OpenSWATH == T) %>% select(protein)),
                    Skyline = unlist(proteinoverlap.r1 %>% filter(Skyline == T) %>% select(protein))
                    ),
                filename = "Figure5C.r1.tiff",
                fontfamily="Arial",
                fill=c("yellow", "red", "green", "cyan"),
                cex=1.5, cat.cex=1.1, cat.dist=0.07
                #,category.names = rep("", 4)
                )

#nrow(proteinoverlap %>% filter(OpenSWATH == T))
```


The overlap between all software tools:

```{r}

l <-  venn.diagram(
                x = list(
                    DIA_Umpire = unlist(proteinoverlap %>% filter(DIAumpire == T) %>% select(protein)),
                    Spectronaut = unlist(proteinoverlap %>% filter(Spectronaut == T) %>% select(protein)),
                    PeakView = unlist(proteinoverlap %>% filter(PeakView == T) %>% select(protein)),
                    OpenSWATH = unlist(proteinoverlap %>% filter(OpenSWATH == T) %>% select(protein)),
                    Skyline = unlist(proteinoverlap %>% filter(Skyline == T) %>% select(protein))
                    ), 
                filename = "ProteinOverlap_allTools.tiff",
                fill=c("blue", "yellow", "red", "green", "cyan"),
                cex=1, cat.cex=1.1, cat.dist=0.07)

l <-  venn.diagram(
                x = list(
                    DIA_Umpire = unlist(proteinoverlap.r1 %>% filter(DIAumpire == T) %>% select(protein)),
                    Spectronaut = unlist(proteinoverlap.r1 %>% filter(Spectronaut == T) %>% select(protein)),
                    PeakView = unlist(proteinoverlap.r1 %>% filter(PeakView == T) %>% select(protein)),
                    OpenSWATH = unlist(proteinoverlap.r1 %>% filter(OpenSWATH == T) %>% select(protein)),
                    Skyline = unlist(proteinoverlap.r1 %>% filter(Skyline == T) %>% select(protein))
                    ), 
                filename = "ProteinOverlap_allTools.r1.tiff",
                fill=c("blue", "yellow", "red", "green", "cyan"),
                cex=1, cat.cex=1.1, cat.dist=0.07)

```


We then check the overlap between proteins identified in all library-based tools and DIA-Umpire: 

```{r overlap}


venn.plot <- venn.diagram(
    x = list(
        library_based = unlist(proteinoverlap %>% filter(Spectronaut == T | PeakView == T |  Skyline == T | OpenSWATH == T) %>% select(protein)),
        DIA_Umpire = unlist(proteinoverlap %>% filter(DIAumpire == T) %>% select(protein))
        ),
    fontfamily="Arial",
    filename = "Figure5D.tiff",
    fill=c("yellow", rgb(0,191/255,196/255)),
    cex=1.5, cat.cex=1.1, cat.dist=0.07,
    category.names = c("Library-based", "DIA-Umpire")  # rep("", 2)
    )

venn.plot <- venn.diagram(
    x = list(
        library_based = unlist(proteinoverlap.r1 %>% filter(Spectronaut == T | PeakView == T |  Skyline == T | OpenSWATH == T) %>% select(protein)),
        DIA_Umpire = unlist(proteinoverlap.r1 %>% filter(DIAumpire == T) %>% select(protein))
        ),
    fontfamily="Arial",
    filename = "Figure5D.r1.tiff",
    fill=c("yellow", rgb(0,191/255,196/255)),
    cex=1.5, cat.cex=1.1, cat.dist=0.07,
    category.names = c("Library-based", "DIA-Umpire")  # rep("", 2)
    )

               
venn.plot <- venn.diagram(
    x = list(
        Spectronaut = unlist(proteinoverlap %>% filter(Spectronaut == T ) %>% select(protein)),
        DIA_Umpire = unlist(proteinoverlap %>% filter(DIAumpire == T) %>% select(protein))
        ),
    filename = "Figure5D_onlySpectronaut.tiff",
    col = "transparent",
    fill = c("red", "blue"),
    alpha = 0.5,
    #label.col = c("darkred", "white", "darkblue", "white",
    # "white", "white", "darkgreen"),
    cex = 1.5,
    fontfamily = "serif",
    fontface = "bold",
    cat.default.pos = "text",
    cat.col = c("darkred", "darkblue"),
    cat.cex = 1.5,
    cat.fontfamily = "serif",
    cat.dist = c(0.03, 0.03),
    cat.pos = c(-20,5)
    )


diaump.only.id <- (proteinoverlap %>% filter(Spectronaut == F & Skyline == F & PeakView == F & OpenSWATH == F & DIAumpire == T))$protein

```

There are only `r length(diaump.only.id)` proteins reported in DIA-Umpire that were not detected in Spectronaut. If we plot the intensities of these proteins (in blue) compared to the intensities of the proteins detected in both tools (in red) we observe that they are in the low range. 


```{r intensities}
diaump.only.id <- (proteinoverlap %>% filter(Spectronaut == F & Skyline == F & PeakView == F & OpenSWATH == F & DIAumpire == T))$protein
diaump <- diaump %>% 
        mutate( onlyDIAump = ifelse(protein %in% diaump.only.id, T, F) ) 

p <- ggplot(diaump, aes(x = log10(meanIntensity)))
p <- p + theme_classic() + theme(text = element_text(size=25))
p <- p + xlab(expression('log'[10]('average Intensity'))) + ylab('number of proteins')
p <- p + geom_histogram(aes(fill = onlyDIAump), position = "dodge" )
p <- p + facet_grid(facets= . ~ specie)

print(p)
ggsave(filename="Figure5E.pdf", plot=p, width=7, height=7)
```

